Radiance - A Web Application Environment

### 2017-03-15 els2017-radiance/paper.tex
Fix some grammar and style problems.

%% FIXME:
%% - !! Elaborate on the provided standard interfaces that Radiance's core provides
%% - Elaborate on mechanisms for required and optional interface dependence
%% - Elaborate on how the interface mechanism is implemented
%% - Elaborate on how to actually implement an interface
%% - Explain what options an adminitrator has to administer the system
%% - Explain difference in URI to W3C URIs
%% - Elaborate on the problem of deployment of single-application programs
%% - Show a real-world example

@@

\section{Introduction}
As the internet evolved, served websites began to evolve and become more and more dynamic. Content is no longer a set of static webpages, but rather automatically updated or even individually composed for the specific user viewing the website. Creating such dynamic websites requires a lot of extra work, much of which is usually done in much the same way for every website. The requested content needs to be retrieved somehow, transformed as needed, and finally assembled into deliverable HTML. \\

In order to handle these common aspects, web frameworks have been created. These frameworks can come in all kinds of sizes, from the very minimal set of a HTML templating engine and a web server as seen in ``micro-frameworks''\cite{microframeworks}, to extensive sets of utilities to handle user interaction and data. \\

Typically these frameworks are constructed with the intent of helping you develop a single web application, which is then deployed and run standalone. However, this can lead to issues when multiple applications should be deployed side-by-side. For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together. Large, ``macro-frameworks'' may avoid the problems introduced by ad-hoc feature implementation in each application, but instead run the risk of introducing too many features that remain unused. \\

Radiance initially grew out of the desire to write web applications that could be run together and in such a way that users would not have to create multiple accounts, track multiple logins, and so forth. Another primary goal of it was to write it in such a way that it would be possible to exchange various parts of the framework without needing to change application code. This would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.

\section{Example Requirements of an Application}
In order to illustrate the reasoning behind the individual components present in the Radiance system, we are going to make use of a simple, imaginary example web application throughout the paper. This application should provide a ``paste service'' wherein users can upload text snippets, to which they can then link to. If they have a registered account, they should be able to manage pastes they have made before, and configure the look of the site to their liking. \\

When talking about an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver. This is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application. \\

In order to write our paste application, we are going to want a database of some sort that stores the pastes, a system to handle user accounts and authentication, a user settings panel, a cache system for performance, an HTTP server, and a template engine. \\

For this application to share as many resources as it can with potential third-party applications residing in the same installation, it needs to rely on the framework to provide all of the above, except for the template engine. Additionally, it needs to have some system that divides up the URL namespace between applications. As the application writer, we cannot have any presumptions about what the final setup will look like-- what kinds of applications will be deployed together, and how the public URLs should resolve.

\section{The Radiance System}
The core of the Radiance system is rather small in terms of actual code. It can provide a plethora of features in the spirit of a macro-framework, but doesn't have to. Most of its features are pluggable, which is done through interfaces. \\

@@

Finally, the setup of a Radiance configuration is handled through so-called ``environments,'' which specify the setup of the individual components and how they can be reached.

@@

This construct defines a new package called \code{cache}, exports the symbols \code{invalidate} and \code{with-caching} from it, and installs stub definitions for the respective function and macro. With this interface specification in place, an application can start writing code against it. In our imaginary paste service, we could now cache the page for a specific paste like so:

\begin{minted}[fontsize=\small]{common-lisp}
(defun paste-page (id)
  (cache:with-caching id NIL
    (render-paste (load-paste id))))
\end{minted}

However, as it currently is, this will obviously not work. The \code{with-caching} macro is not implemented with any useful functionality. As the writer of the paste application, we don't need to know the specifics of how the caching is implemented, though. All we need to do is tell Radiance that we need this interface. This can be done by adding the interface as a dependency to our application's ASDF\cite{asdf} system definition.

\begin{minted}[fontsize=\small]{common-lisp}
(asdf:defsystem paste-service
  ...
  :depends-on (...
               (:interface :cache)))
\end{minted}

Radiance extends ASDF's dependency resolution mechanism to allow for this kind of declaration. When the \\\code{paste-service} system is now loaded, Radiance will notice the interface dependency and resolve it to an actual system that implements the desired cache functionality. This resolution is defined in the currently active environment, and can thus be configured by the administrator of the Radiance installation. \\

For completeness, let's look at an implementation of this cache interface. The implementation must be loadable through an ASDF system, so that the above dependency resolution can be done. The actual implementation mechanism is handed to us by Lisp allowing redefinition.

@@

Using direct overwriting of definitions means that all applications must use the same implementation. However, usually this is the intended effect, as we want to maximise the sharing between applications. If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on. This approach does bring for some great benefits however. For one, it allows the interfaces to be as efficient as possible, as there is no intermediate layer of redirection. It also allows us to expose any kind of definition, including custom, user-defined ones. \\

Radiance provides standard interface definitions for all of the components we require for the paste application, and more. A full list of the interfaces and their descriptions is available in the Radiance documentation\footnote{\url{https://github.com/Shirakumo/radiance/\#interface}}. Thus, Radiance can be used like a macro-framework, but does not load on any features unless specifically required. Additionally, any of the features can be exchanged by a system administrator for ones that more closely match their requirements, without having to change anything about the application code.\\

Finally, Radiance provides a system by which functionality can be provided if a specific interface should be loaded, without explicitly having to depend on it. This is useful to model something like the user settings mentioned in our example application. An administrator might not always want to provide an administration or settings panel. To make this dependency optional, we can defer the compilation and evaluation of our relevant logic to a later point, after an implementation has been loaded. For an imaginary \code{user-settings} interface, this might look like the following:

@@

Since all the symbols are already provided by the interface definition, there are no problems when the reader parses the source. The forms can thus simply be saved away for evaluation once Radiance notices that an implementation of the interface in question has been loaded. \\

Ultimately, interfaces are a form of compromise between providing all possible features at once, and almost no features at all. The usefulness of an interface heavily depends on its specification, and in order for implementations to be really exchangeable without modifying the application code, each implementation and application must strictly adhere to the specification. Thus interfaces are very restricting, and often cannot expose the full power of an underlying system, to allow a variety of different backing implementations.

\subsection{Routes}
In order to allow the administrator to change where pages are accessible from, an application cannot hardcode its resource locations and the links in its templates. This is doubly important, when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another. \\

@@

This separation allows application programmers to model their pages from a perspective that looks much more like the external view. Instead of dealing with named tags, they deal in concrete internal URLs that in a development setup often have a relatively direct one-to-one mapping to the external URLs. This makes it easier to visualise and think about the application structure. From the administrator's point of view this setup is more convenient, as they too simply need to think in terms of translations, rather than having to adapt potentially hundreds of different URL tags manually. \\

@@

Since Radiance has full control over the organisation of the internal representation, it can make strict demands as to how applications need to structure their URLs in order for the translation to work correctly. Specifically, it requires each application to put all of its pages on a domain that is named after their application. \\

In our sample application, we might write the view page something like this, where \texttt{paste/view} is the internal URL that the view page is reachable on.

\begin{minted}[fontsize=\small]{common-lisp}
(define-page view "paste/view" ()
  (render-template (template "view.html")
                   (get-paste (get-var "paste-id"))))
\end{minted}

If, as an administrator of an installation of this paste application, we now wanted to reach this page through the URL \texttt{www.example.com/paste/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{paste} folder, and put that as the URL's domain instead. We can achieve this relatively easily with the following definition.

\begin{minted}[fontsize=\small]{common-lisp}
(define-route paste :mapping (uri)
  (when (begins-with "paste/" (path uri))
    (setf (domains uri) '("paste")
          (path uri) (subseq (path uri) 6))))
\end{minted}

@@

\begin{minted}[fontsize=\small]{common-lisp}
(define-string-route paste :mapping
  "/paste/(.*)" "paste/\\1")
\end{minted}

@@

With the mapping alone the case is not yet solved, however. When a page is emitted that contains links, the links too must be translated, but in the opposite way. Since routes can be arbitrarily complex, it is not possible for the system to figure out a reversal route automatically. The logic of a reversal route will usually be much the same as it was for the mapping route and should thus not be a problem to figure out, though. \\

@@

\begin{minted}[fontsize=\small]{html}
<a @href="paste/list">Public Paste List</a>
<form method="post">
  <textarea name="text"></textarea>
  <input type="submit" @formaction="paste/submit" />
</form>
\end{minted}

@@

In our example application we might want to leave it up to the administrator whether to allow anonymous users to post a paste or not. We can set up a default value like this.

\begin{minted}[fontsize=\small]{common-lisp}
(define-trigger startup ()
  (defaulted-config T :anonymous-pastes))
\end{minted}

@@

While Radiance does provide default implementations for all of the interfaces, it is likely that some of them are not usable for a production setting. In order to change, say, the database interface's implementation, we would then have to modify the Radiance core's configuration file. We can either modify the file directly, which can be found through \code{(mconfig-pathname :radiance-core)} like so:

@@

.. or instead use the programmatical way.

@@

\begin{minted}[fontsize=\small]{common-lisp}
((:routes (paste :mapping "/paste/(.*)" "paste/\\1"))
 ...)
\end{minted}

@@

\section{Further Work}
Currently, while it is possible to dynamically load applications, it is not possible to dynamically unload them. Doing so is troublesome, as an application's system might potentially modify any other part of the Lisp process. However, if the changes that can be rolled back is constrained in some sense, it should be possible to provide this kind of feature. \\

Radiance also does not allow you to change the implementation of an interface on the fly. This has much the same problems as the previous issue. Implementation switching is even more problematic though, since the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect. Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to migrate objects between interface implementations, or somehow invalidate the old instances if they are retained in another part of the application. \\

>>>>

@@

\section{Introduction}
As the internet evolved, served websites began to evolve and became more and more dynamic. Content is no longer a set of static webpages, but rather automatically updated, or even individually composed for the specific user viewing the website. Creating such dynamic websites requires a lot of extra work, much of which is usually done in much the same way for every website. The requested content needs to be retrieved somehow, transformed as needed, and finally assembled into deliverable HTML. \\

In order to handle these common aspects, web frameworks have been created. These frameworks can come in all kinds of sizes, from the very minimal set of an HTML templating engine and a web server as seen in ``micro-frameworks''\cite{microframeworks}, to extensive sets of utilities to handle user interaction and data of all kinds. \\

Typically these frameworks are constructed with the intent of helping you develop a single web application, which is then deployed and run standalone. However, this can lead to issues when multiple applications should be deployed side-by-side. For example, if the framework does not provide explicit support for a particular feature such as user accounts, the two applications will likely have implemented ad-hoc solutions of their own, which are incompatible with each other and thus can't be trivially merged together. Large, ``macro-frameworks'' may avoid the problems introduced by ad-hoc feature implementation, but instead run the risk of introducing too many features that remain unused. \\

Radiance initially grew out of the desire to write web applications that could be run together in such a way, that users would not have to create multiple accounts, track multiple logins, and so forth. Another primary goal of it was to write it in such a way, that it would be possible to exchange various parts of the framework without needing to change application code. This would allow flexibility on the side of the administrator, to pick and choose the parts that best fit their environment.

\section{Example Requirements of an Application}
In order to illustrate the reasoning behind the individual components present in the Radiance system, we are going to make use of a simple, imaginary example web application throughout the paper. This application should provide a ``blog platform,'' wherein users can create text posts to which they can link their friends. They should also be able to edit the posts again at a later point and customise the look and feel to their preference. \\

When talking about such an application in the context of Radiance, we think of it as a form of library that is then loaded together with Radiance to form a full webserver. This is in contrast to the deployment models of many other frameworks, where the ``library'' is deployed bundled tightly together with the framework as a single application. \\

In order to write our blog application, we are going to want a database that stores the posts, a system to handle user accounts and authentication, a cache system for performance, an HTTP server, and a template engine. \\

Imagine now, if you will, that an administrator also wanted to run a forum on the same host as well, perhaps in order to allow people to discuss and engage with each other about the various posts on the main blog. Requiring users to maintain a login for each service would be annoying, to say the least. Furthermore, as an administrator, we would like to make sure that as much data is kept in the same place as possible, and arranged in a similar way, to ease maintenance. \\

For our blog to share as many resources as it can with potential third-party applications residing in the same installation, it needs to rely on the framework to provide all of our required features, except perhaps for the template engine. Additionally, it needs to have some system that divides up the URL namespace between applications. As the application writer, we cannot have any presumptions about what the final setup will look like-- what kinds of applications will be deployed together, and how the public URLs should resolve. \\

\section{The Radiance System}
The core of the Radiance system is rather small. Despite this, it can provide a plethora of features in the spirit of a macro-framework, if needed. Most of its features are pluggable, which is done through interfaces. \\

@@

Finally, the management of a Radiance installation is handled through so-called ``environments,'' which specify the configuration of the individual components.

@@

This construct defines a new package called \code{cache}, exports the symbols \code{invalidate} and \code{with-caching} from it, and installs stub definitions for the respective function and macro. With this interface specification in place, an application can start writing code against it. In our imaginary blog service, we could now cache the page for a specific post like so:

\begin{minted}[fontsize=\small]{common-lisp}
(defun post-page (id)
  (cache:with-caching id NIL
    (render-post (load-post id))))
\end{minted}

However, as it currently is, this will obviously not work. The \code{with-caching} macro is not implemented with any useful functionality. As the writer of the blog application, we don't need to know the specifics of how the caching is implemented, though. All we need to do is tell Radiance that we need this interface. This can be done by adding the interface as a dependency to our application's ASDF\cite{asdf} system definition.

\begin{minted}[fontsize=\small]{common-lisp}
(asdf:defsystem blog-service
  ...
  :depends-on (...
               (:interface :cache)))
\end{minted}

Radiance extends ASDF's dependency resolution mechanism to allow for this kind of declaration. When the \\\code{blog-service} system is now loaded, Radiance will notice the interface dependency and resolve it to an actual system that implements the desired cache functionality. This resolution is defined in the currently active environment, and can thus be configured by the administrator of the Radiance installation. \\

For completeness, let's look at an implementation of this cache interface. The implementation must be loadable through an ASDF system, so that the above dependency resolution can be done. The actual implementation mechanism is handed to us for free, as Lisp allows redefinition.

@@

Using direct overwriting of definitions means that all applications must use the same implementation. However, usually this is the intended effect, as we want to maximise the sharing between them. If an implementation should have special needs, it can always bypass the interfaces and make direct use of whatever it might depend on. This interfaces approach does bring some benefits, however. For one, it allows the implementations to be as efficient as possible, as there is no intermediate layer of redirection. \\

Radiance provides standard interface definitions for all of the components we require for the blog application, and more. A full list of the interfaces and their descriptions is available in the Radiance documentation\footnote{\url{https://github.com/Shirakumo/radiance/\#interface}}. Thus, Radiance can be used like a macro-framework, but does not load on any features unless specifically required. Additionally, any of the implementations can be exchanged by a system administrator for ones that more closely match their requirements, without having to change anything about the application code.\\

Finally, Radiance provides a system that allows to optionally depend on an interface. This is useful to model something like the user preferences mentioned in our example application. An administrator might not always want to provide an administration or settings panel. To make this dependency optional, we can defer the compilation and evaluation of our relevant logic to a later point, after an implementation has been loaded. For an imaginary \code{user-settings} interface, this might look like the following:

@@

Since all the symbols are already provided by the interface definition, there are no problems when the reader parses the source. The forms can thus simply be saved away to be evaluated once Radiance notices that an implementation of the interface in question has been loaded. \\

Ultimately, interfaces are a form of compromise between providing all possible features at once, and almost no features at all. The usefulness of an interface heavily depends on its specification, and for implementations to be really exchangeable without modifying the application code, each implementation and application must strictly adhere to the specification.

\subsection{Routes}
In order to allow the administrator to change where pages are accessible from, an application cannot hard-code its resource locations and the links in its templates. This is doubly important when it comes to housing multiple applications in one, as the system needs to be set up in such a way that the applications do not clash with each other or potentially confuse pages of one another. \\

@@

This separation allows application programmers to model their pages from a perspective that looks much more like the external view. Instead of dealing with named tags, they deal in concrete internal URLs that, in a development setup, often have a relatively direct one-to-one mapping to the external URLs. This makes it easier to visualise and think about the application structure. From the administrator's point of view this setup is more convenient too, as they simply need to think in terms of translations, rather than having to adapt potentially hundreds of different URL tags manually. \\

@@

Since Radiance has full control over the organisation of the internal representation, it can make strict demands as to how applications need to structure their URLs. Specifically, it requires each application to put all of its pages on a domain that is named after their application. \\

In our sample application, we might write the view page something like this, where \texttt{blog/view} is the internal URL that the view page is reachable on.

\begin{minted}[fontsize=\small]{common-lisp}
(define-page view "blog/view" ()
  (render-template (template "view.html")
                   (get-post (get-var "post-id"))))
\end{minted}

If, as an administrator of an installation of this blog application, we now wanted to reach this page through the URL \texttt{www.example.com/blog/view}, the mapping route functions would have to strip away the \texttt{www.example.com} domain, recognise the \texttt{blog} folder, and put that as the URL's domain instead. We can achieve this relatively easily with the following definition.

\begin{minted}[fontsize=\small]{common-lisp}
(define-route blog :mapping (uri)
  (when (begins-with "blog/" (path uri))
    (setf (domains uri) '("blog")
          (path uri) (subseq (path uri) 5))))
\end{minted}

@@

\begin{minted}[fontsize=\small]{common-lisp}
(define-string-route blog :mapping
  "/blog/(.*)" "blog/\\1")
\end{minted}

@@

With the mapping alone the case is not yet solved, though. When a page is emitted that contains links, the links too must be translated, but in the opposite way. Unfortunately, because routes can be arbitrarily complex, it is not possible for the system to figure out a reversal route automatically. The logic of a reversal route will usually be much the same as it was for the mapping route and should thus not be a problem to figure out, though. \\

@@

\begin{minted}[fontsize=\small]{html}
<a @href="blog/list">Latest Blog Posts</a>
<form method="post">
  <textarea name="text"></textarea>
  <input type="submit" @formaction="blog/submit" />
</form>
\end{minted}

@@

In our example application we might want to give the administrator the ability to define a global blog title. We could provide a default value for it like this:

\begin{minted}[fontsize=\small]{common-lisp}
(define-trigger startup ()
  (defaulted-config "Irradiant Blogs" :title))
\end{minted}

@@

While Radiance does provide default implementations for all of its interfaces, it is likely that some of them are not usable for a production setting. In order to change, say, the database interface's implementation, we would then have to modify the Radiance core's configuration file. We can either modify the file directly, which can be found through \code{(mconfig-pathname :radiance-core)}, like so:

@@

.. or instead use the programmatical way, like so:

@@

\begin{minted}[fontsize=\small]{common-lisp}
((:routes (blog :mapping "/blog/(.*)" "blog/\\1"))
 ...)
\end{minted}

@@

\section{Further Work}
Currently, while it is possible to dynamically load applications, it is not possible to dynamically unload them. Doing so is troublesome, as an application's system might potentially modify any other part of the Lisp process. However, if there is a constraint on what kind of changes can be rolled back, it should be possible to provide a usable form of this kind of feature. \\

Radiance also does not allow you to change the implementation of an interface on the fly. This has much the same problems as the previous issue. In addition, though, the system needs to ensure that all dependant applications, including optionally dependant ones, are reloaded to make macro redefinitions take effect. Furthermore, since some interfaces expose classes and instances thereof, the system would either have to be able to migrate objects between interface implementations, or somehow invalidate the old instances if they are retained in another part of the application. \\
